\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-06-23 paper.qmd
added lines corresponding to sub-optimal bin sizes to MSE plot

hex_grid_coloured_scurve1 <- ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts_s_curve1, 
    aes(x = x, y = y,
      group = hex_poly_id, 
      fill = std_counts), color = "grey70", linewidth=0.2) +
  geom_point(data = umap_scurve_scaled,
             aes(x = UMAP1, y = UMAP2),
             alpha = 0.3,
             size = 0.5) +
  scale_fill_viridis_c(direction = -1, 
    na.value = "#ffffff", option = "C") +
  interior_annotation("a") +
  xlim(sc_xlims) + ylim(sc_ylims)
```

```{r}
## hexagon binning to have regular hexagons
hb_obj_scurve2 <- hex_binning(
  data = umap_scurve_scaled, 
  bin1 = 12, 
  r2 = r2,
  q = 0.07)

@@

hex_grid_coloured_scurve2 <- ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts_s_curve2, 
    aes(x = x, y = y, 
        group = hex_poly_id, 
        fill = std_counts), 
        color = "grey70", 
        linewidth=0.2) +
  geom_point(data = umap_scurve_scaled,
           aes(x = UMAP1, y = UMAP2),
           alpha = 0.3,
           size = 0.5) +
  scale_fill_viridis_c(direction = -1, 
    na.value = "#ffffff", option = "C") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("b")

@@

hex_grid_coloured_scurve3 <-  ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts_s_curve3, 
    aes(x = x, y = y, 
        group = hex_poly_id, 
        fill = std_counts), 
        color = "grey70", linewidth=0.2) +
  geom_point(data = umap_scurve_scaled,
           aes(x = UMAP1, y = UMAP2),
           alpha = 0.3,
           size = 0.5) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("c")

@@

```{r}
#| echo: false
#| label: fig-bins-scurve
#| fig-pos: H
#| fig-cap: "Hexbin plots from different number of bins for the UMAP applied to S-curve data: (a) $b = 91$ (7, 13), (b) $b = 264$ (12, 22), and (c) $b = 364$ (14, 26). The hexbins are colored based on the density of points, with darker colors indicating higher density and yellow color representing lower density within each bin. What is the number of bins that would be effective?"
#| fig-width: 6
#| fig-height: 3

@@

  bin2 <- calc_bins_y(bin1 = xbins, r2 = r2, q = 0.07)$bin2

  scurve_model <- fit_highd_model(
    training_data = training_data_scurve,
    emb_df = umap_scurve_scaled,
    bin1 = xbins,
    r2 = r2,
    q = 0.07, 
    is_bin_centroid = TRUE,
    is_rm_lwd_hex = FALSE,
    col_start_highd = "x"
  )

@@

mse_scurve_b <- ggplot(error_scurve, 
                     aes(x = b, 
                         y = log(MSE))) + 
  geom_point(size = 1) +
  geom_line() + 
  geom_vline(xintercept = 264, linetype="solid", 
             color = "black", linewidth=0.5, alpha = 0.5) +
  ylab("log(MSE)") +
  xlab("b") +
  ggtitle("(a)") +
  theme(aspect.ratio = 0.75,
        axis.text.x = element_text(size = 5),
        axis.text.y = element_text(size = 5),
        axis.title.x = element_text(size = 5),
        axis.title.y = element_text(size = 5, 
            angle=90),
        title = element_text(size = 36))
```

@@

```{r}
## To initialize effective bins along x
effective_bin1_scurve <- 12

effective_bin2_scurve <- calc_bins_y(
  bin1 = effective_bin1_scurve, 
  r2 = r2, 
  q = 0.07)$bin2

scurve_model <- fit_highd_model(
  training_data = training_data_scurve,
  emb_df = umap_scurve_scaled,
  bin1 = effective_bin1_scurve,
  r2 = r2,
  q = 0.07,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "x"
)

@@

```{r}
cell_count_scurve <- ggplot(df_bin_centroids_scurve, 
                            aes(x = reorder(as.factor(hexID), 
                                            -std_counts), 
                                y = std_counts)) + 
  geom_quasirandom(
    size = 0.3,
    alpha = 0.5
  ) + 
  geom_hline(yintercept = benchmark1, color = "black", linewidth=0.5, alpha = 0.5) +
  xlab("hexagon id (re-ordered)") + 
  ylab("standardized bin count") +
  ggtitle("(b)") +
  theme(aspect.ratio = 0.75,
        axis.title.x = element_text(size = 5),
        axis.title.y = element_text(size = 5, angle = 90),
        axis.text.x = element_blank(),
        axis.text.y = element_text(size = 5),
        title = element_text(size = 36))
```

@@

@@

mse_scurve_lwd <- ggplot(error_rm_scurve, 
                     aes(x = benchmark_rm_lwd, 
                         y = log(MSE))) + 
  geom_point(size = 1) +
  geom_line() + 
  geom_vline(xintercept = benchmark1, linetype="solid", 
             color = "black", linewidth=0.5, alpha = 0.5) +
  scale_x_continuous("standardized bin count", 
         transform = "reverse") + 
  ylab("log(MSE)") +
  ggtitle("(c)") +
  theme(aspect.ratio = 0.75,
        axis.text.x = element_text(size = 5),
        axis.text.y = element_text(size = 5),
        axis.title.x = element_text(size = 5),
        axis.title.y = element_text(size = 5, 
            angle=90),
        title = element_text(size = 36))
```

@@

<!--Using first quantile as the benchmark value-->
```{r}
hex_grid_with_counts_s_curve2 <- full_join(
  hex_grid2, 
  df_bin_centroids_scurve |> select(hexID, std_counts), 
  by = c("hex_poly_id" = "hexID"))

hex_grid_with_low_counts <- full_join(hex_grid2, 
    df_bin_centroids_scurve, 
    by = c("hex_poly_id" = "hexID")) |>
  filter(!is.na(std_counts)) |>
  mutate(type = if_else(std_counts <= benchmark1, "low", "high")) |>
  filter(type == "low")

hex_grid_rm_lwd_scurve1 <-  ggplot(
  data = hex_grid_with_counts_s_curve2, 
  aes(x = x, y = y)) + 
  geom_polygon(aes(group = hex_poly_id, fill = std_counts), 
            color = "grey70", linewidth=0.2) +
  geom_text(data = hex_grid_with_low_counts, 
            color = "black", 
            aes(x = c_x, y = c_y, group = hex_poly_id, label = "X", size = 1)) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("a")
```

@@

hex_grid_rm_lwd_scurve2 <-  ggplot(
  data = hex_grid_with_counts_s_curve2, 
  aes(x = x, y = y)) + 
  geom_polygon(aes(group = hex_poly_id, fill = std_counts), 
            color = "grey70", linewidth=0.2) +
  geom_text(data = hex_grid_with_low_counts_neighbor, 
            color = "black", 
            aes(x = c_x, y = c_y, group = hex_poly_id, label = "X", size = 1)) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("b")
```

@@

## To draw the distance distribution
distance_scurve$group <- "1"
distance_scurve_plot <- ggplot(distance_scurve, 
                        aes(x = group, 
                            y = distance)) + 
  geom_quasirandom(
    size = 0.5, 
    alpha = 0.3
  ) + 
  ylim(0, max(unlist(distance_scurve$distance))+ 0.5) + 
  coord_flip() + 
  geom_hline(yintercept = benchmark_scurve,
     linetype="solid", 
     color = "black", linewidth=0.5, alpha = 0.5) +
  ylab(expression(d^{(2)})) +
  ggtitle("(d)") +
  theme(aspect.ratio = 0.75,
        axis.text.x = element_text(size = 5),
        axis.title.x = element_text(size = 5),
        title = element_text(size = 36))
```

```{r}
#| echo: false
#| fig-cap: "Various plots to help assess best number of bins, low density bin and large edge removal. Both (a) and (c) show MSE, against number of bins and standardised count. A good benchmark value for these parameters is when the MSE drops and then flattens out."
# Distribution of $2\\text{-}D$ Euclidean distances between bin centroids of the triangular mesh generated with S-curve UMAP data.
#| label: fig-dist-scurve
#| out-width: 100%
#| fig-width: 8
#| fig-height: 6
#| fig-pos: H

>>>>

hex_grid_coloured_scurve1 <- ggplot() + 
  geom_point(data = umap_scurve_scaled, 
             aes(x = UMAP1, y = UMAP2), 
             alpha = 0.5,
             size = 0.5) + 
  geom_polygon(
    data = hex_grid_with_counts_s_curve1, 
    aes(x = x, y = y,
      group = hex_poly_id, 
      fill = std_counts), color = "grey70", linewidth=0.2) +
  scale_fill_viridis_c(direction = -1, 
    na.value = "#ffffff", option = "C") +
  interior_annotation("a") +
  xlim(sc_xlims) + ylim(sc_ylims)
```

```{r}
## hexagon binning to have regular hexagons
hb_obj_scurve2 <- hex_binning(
  data = umap_scurve_scaled, 
  bin1 = 10, 
  r2 = r2,
  q = 0.05)

@@

hex_grid_coloured_scurve2 <- ggplot() + 
  geom_point(data = umap_scurve_scaled, 
             aes(x = UMAP1, y = UMAP2), 
             alpha = 0.5,
             size = 0.5) + 
  geom_polygon(
    data = hex_grid_with_counts_s_curve2, 
    aes(x = x, y = y, 
        group = hex_poly_id, 
        fill = std_counts), 
        color = "grey70", 
        linewidth=0.2) +
  scale_fill_viridis_c(direction = -1, 
    na.value = "#ffffff", option = "C") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("b")

@@

hex_grid_coloured_scurve3 <-  ggplot() + 
  geom_point(data = umap_scurve_scaled, 
             aes(x = UMAP1, y = UMAP2), 
             alpha = 0.5,
             size = 0.5) + 
  geom_polygon(
    data = hex_grid_with_counts_s_curve3, 
    aes(x = x, y = y, 
        group = hex_poly_id, 
        fill = std_counts), 
        color = "grey70", linewidth=0.2) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("c")

@@

```{r}
#| echo: false
#| label: fig-bins-scurve
#| fig-pos: H
#| fig-cap: "Hexbin density plots of UMAP layout of the S-curve data, using three different bin inputs: (a) $b = 91$ (7, 13), (b) $b = 190$ (10, 19), and (c) $b = 364$ (14, 26). Color indicates standardized counts, dark indicating high count and light indicates low count. At the smallest bin size the data segregates into two separate groups, suggesting this is too many bins. Using the MSE of the model fit in $p-D$ helps decide on a useful choice of number of bins."
#| fig-width: 6
#| fig-height: 3

@@

  bin2 <- calc_bins_y(bin1 = xbins, r2 = r2)$bin2

  scurve_model <- fit_highd_model(
    training_data = training_data_scurve,
    emb_df = umap_scurve_scaled,
    bin1 = xbins,
    r2 = r2,
    is_bin_centroid = TRUE,
    is_rm_lwd_hex = FALSE,
    col_start_highd = "x"
  )

@@

mse_scurve_b <- ggplot(error_scurve, 
                     aes(x = b, 
                         y = log(MSE))) + 
  geom_point(size = 1) +
  geom_line() + 
  geom_vline(xintercept = 190, linetype="solid", 
             color = "black", linewidth=0.8, alpha = 0.5) +
  geom_vline(xintercept = 91, linetype=2, 
             color = "black", linewidth=0.8, alpha = 0.5) +
  geom_vline(xintercept = 364, linetype=2, 
             color = "black", linewidth=0.8, alpha = 0.5) +
  ylab("log(MSE)") +
  xlab("b") +
  ggtitle("(a)") +
  theme(aspect.ratio = 0.75,
        axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10, 
            angle=90),
        title = element_text(size = 36))
```

@@

```{r}
## To initialize effective bins along x
effective_bin1_scurve <- 10

effective_bin2_scurve <- calc_bins_y(bin1 = effective_bin1_scurve, r2 = r2)$bin2

scurve_model <- fit_highd_model(
  training_data = training_data_scurve,
  emb_df = umap_scurve_scaled,
  bin1 = effective_bin1_scurve,
  r2 = r2,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "x"
)

@@

```{r}
cell_count_scurve <- ggplot(df_bin_centroids_scurve, 
                            aes(x = reorder(as.factor(hexID), 
                                            -std_counts), 
                                y = std_counts)) + 
  geom_quasirandom() + 
  geom_hline(yintercept = benchmark1, color = "black", linewidth=0.8, alpha = 0.5) +
  xlab("hexagon id (re-ordered)") + 
  ylab("standardized bin count") +
  ggtitle("(b)") +
  theme(aspect.ratio = 0.75,
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10, angle = 90),
        #axis.text.x = element_text(size = 7, angle = 90),
        axis.text.x = element_blank(),
        axis.text.y = element_text(size = 10),
        title = element_text(size = 36))
```

@@

scurve_model <- fit_highd_model(
  training_data = training_data_scurve,
  emb_df = umap_scurve_scaled,
  bin1 = effective_bin1_scurve,
  r2 = r2,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "x"
)

@@

mse_scurve_lwd <- ggplot(error_rm_scurve, 
                     aes(x = benchmark_rm_lwd, 
                         y = log(MSE))) + 
  geom_point(size = 1) +
  geom_line() + 
  geom_vline(xintercept = benchmark1, linetype="solid", 
             color = "black", linewidth=0.8, alpha = 0.5) +
  geom_text(data = benchmark_label_df, 
            aes(x=x, 
                label= label, 
                y=y), 
            colour="black", 
            angle=90) +
  scale_x_continuous("standardized bin count", 
         transform = "reverse") + 
  ylab("log(MSE)") +
  ggtitle("(c)") +
  theme(aspect.ratio = 0.75,
        axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10, 
            angle=90),
        title = element_text(size = 36))
```

@@

<!--Using first quantile as the benchmark value-->
```{r}
hex_grid_with_low_counts <- full_join(hex_grid2, 
    df_bin_centroids_scurve, 
    by = c("hex_poly_id" = "hexID")) |>
  filter(!is.na(std_counts)) |>
  mutate(type = if_else(std_counts <= benchmark1, "low", "high")) |>
  filter(type == "low")

hex_grid_rm_lwd_scurve1 <-  ggplot(
  data = hex_grid_with_counts_s_curve2, 
  aes(x = x, y = y)) + 
  geom_polygon(color = "black", aes(group = hex_poly_id, fill = std_counts)) +
  geom_text(data = hex_grid_with_low_counts, 
            color = "black", 
            aes(x = c_x, y = c_y, group = hex_poly_id, label = "X", size = 1)) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("a")
```

@@

hex_grid_rm_lwd_scurve2 <-  ggplot(
  data = hex_grid_with_counts_s_curve2, 
  aes(x = x, y = y)) + 
  geom_polygon(color = "black", aes(group = hex_poly_id, fill = std_counts)) +
  geom_text(data = hex_grid_with_low_counts_neighbor, 
            color = "black", 
            aes(x = c_x, y = c_y, group = hex_poly_id, label = "X", size = 1)) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("b")
```

@@

## To draw the distance distribution
distance_scurve$group <- "1"
distance_scurve_plot <- ggplot(distance_scurve, 
                        aes(x = group, 
                            y = distance)) + 
  geom_quasirandom() + 
  ylim(0, max(unlist(distance_scurve$distance))+ 0.5) + 
  coord_flip() + 
  geom_hline(yintercept = benchmark_scurve,
     linetype="solid", 
     color = "black", linewidth=0.8, alpha = 0.5) +
  geom_text(data = benchmark_label_df, 
            aes(x=x, 
                label= label, 
                y=y), 
            colour="black", 
            angle=90) +
  ylab(expression(d^{(2)})) +
  ggtitle("(d)") +
  theme(aspect.ratio = 0.75,
        axis.text.x = element_text(size = 10),
        axis.title.x = element_text(size = 10),
        title = element_text(size = 36))
```

```{r}
#| echo: false
#| fig-cap: "Various plots to help assess best number of bins, low density bin and large edge removal. Both (a) and (c) show MSE, against number of bins and standardised count. A good benchmark value for these parameters is when the MSE drops and then flattens out."
# Distribution of $2\\text{-}D$ Euclidean distances between bin centroids of the triangular mesh generated with S-curve UMAP data.
#| label: fig-dist-scurve
#| out-width: 100%
#| fig-width: 10
#| fig-height: 8
#| fig-pos: H

### 2024-06-23 paper.tex
Update with xbins 12
added lines corresponding to sub-optimal bin sizes to MSE plot

\caption{\label{fig-bins-scurve}Hexbin plots from different number of
bins for the UMAP applied to S-curve data: (a) \(b = 91\) (7, 13), (b)
\(b = 190\) (10, 19), and (c) \(b = 364\) (14, 26). The hexbins are
colored based on the density of points, with darker colors indicating
higher density and yellow color representing lower density within each
bin. What is the number of bins that would be effective?}

>>>>

\caption{\label{fig-bins-scurve}Hexbin density plots of UMAP layout of
the S-curve data, using three different bin inputs: (a) \(b = 91\) (7,
13), (b) \(b = 190\) (10, 19), and (c) \(b = 364\) (14, 26). Color
indicates standardized counts, dark indicating high count and light
indicates low count. At the smallest bin size the data segregates into
two separate groups, suggesting this is too many bins. Using the MSE of
the model fit in \(p-D\) helps decide on a useful choice of number of
bins.}
